---
title: "Reproducible Research Project 2"
author: "Antonio Quintieri"
date: "January 19, 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Synopsis
National Weather Service Storm data set is read and analyzed to determine the types of storm events that are the most harmful to population health as well as the highest economic consequences. 
The analysis results show that the biggest source of property damage in the United States comes from Tornados, followed by Floods and Flash Flood. The highest contributor to crop damage comes from hail and is followed by flood and flash floods. The highest source of injuries and fatalities from weather events both come from tornados. 
The second highest fatality source comes from flood sources and is followed by lightning then wind. The second largest source of injuries comes from wind followed by lightning sources.  

## Data Processing and open libraries 
The following is the reading and initialization of the dataset:
```{r}
    #Read and initialize library and data
    library(ggplot2)
    library(grid)
    library(gridExtra)
    library(dplyr)
    library(data.table)
    
    #Cache and reads data if available
      
    storm <- read.csv("repdata_data_StormData.csv", stringsAsFactors=FALSE)
head(storm)
    # Extracting the required data for health and economic impact analysis against weather

event <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", 
           "CROPDMGEXP")
data <- storm[event]
```
##Finding crop damage

Crop damage exponents for each level was listed out and assigned those values for the crop exponent data. Invalid data was excluded by assigning the value as '0'. Then crop damage value was calculated by multiplying the crop damage and crop exponent value.The code for this process was listed below
```{r}
# Finding the property damage exponent and levels
unique(data$PROPDMGEXP)
```
```{r}
# Assigning values for the property exponent data 
data$PROPEXP[data$PROPDMGEXP == "K"] <- 1000
data$PROPEXP[data$PROPDMGEXP == "M"] <- 1e+06
data$PROPEXP[data$PROPDMGEXP == ""] <- 1
data$PROPEXP[data$PROPDMGEXP == "B"] <- 1e+09
data$PROPEXP[data$PROPDMGEXP == "m"] <- 1e+06
data$PROPEXP[data$PROPDMGEXP == "0"] <- 1
data$PROPEXP[data$PROPDMGEXP == "5"] <- 1e+05
data$PROPEXP[data$PROPDMGEXP == "6"] <- 1e+06
data$PROPEXP[data$PROPDMGEXP == "4"] <- 10000
data$PROPEXP[data$PROPDMGEXP == "2"] <- 100
data$PROPEXP[data$PROPDMGEXP == "3"] <- 1000
data$PROPEXP[data$PROPDMGEXP == "h"] <- 100
data$PROPEXP[data$PROPDMGEXP == "7"] <- 1e+07
data$PROPEXP[data$PROPDMGEXP == "H"] <- 100
data$PROPEXP[data$PROPDMGEXP == "1"] <- 10
data$PROPEXP[data$PROPDMGEXP == "8"] <- 1e+08
# Assigning '0' to invalid exponent data
data$PROPEXP[data$PROPDMGEXP == "+"] <- 0
data$PROPEXP[data$PROPDMGEXP == "-"] <- 0
data$PROPEXP[data$PROPDMGEXP == "?"] <- 0
# Calculating the property damage value
data$PROPDMGVAL <- data$PROPDMG * data$PROPEXP
```
```{r}
# Exploring the crop exponent data
unique(data$CROPDMGEXP)
# Assigning values for the crop exponent data 
data$CROPEXP[data$CROPDMGEXP == "M"] <- 1e+06
data$CROPEXP[data$CROPDMGEXP == "K"] <- 1000
data$CROPEXP[data$CROPDMGEXP == "m"] <- 1e+06
data$CROPEXP[data$CROPDMGEXP == "B"] <- 1e+09
data$CROPEXP[data$CROPDMGEXP == "0"] <- 1
data$CROPEXP[data$CROPDMGEXP == "k"] <- 1000
data$CROPEXP[data$CROPDMGEXP == "2"] <- 100
data$CROPEXP[data$CROPDMGEXP == ""] <- 1
# Assigning '0' to invalid exponent data
data$CROPEXP[data$CROPDMGEXP == "?"] <- 0
# calculating the crop damage value
data$CROPDMGVAL <- data$CROPDMG * data$CROPEXP
```
##Finding totals of each incident by event type.

It was observed that " most harmful to population health" events are fatalities and injuries.So,only those events with fatalities and injuries were selecetd.

It was observed that " most harmful to econamic problem"" events are Property and crop damages.So,only those events with property and crop damage were selecetd.

Then for each incident (Fatalities,Injuries, Property damage and Crop damage), the total values were estimated. Code for which is as follows.
```{r}
# Totalling the data by event
fatal <- aggregate(FATALITIES ~ EVTYPE, data, FUN = sum)
injury <- aggregate(INJURIES ~ EVTYPE, data, FUN = sum)
propdmg <- aggregate(PROPDMGVAL ~ EVTYPE, data, FUN = sum)
cropdmg <- aggregate(CROPDMGVAL ~ EVTYPE, data, FUN = sum)
```
##Plotting events with highest fatalities and highest injuries.

Highest fatalities and highest injuries for Top 8 events were calculated. For better understanding and comparision these values were plotted as follows.
```{r}
# Listing  events with highest fatalities
fatal8 <- fatal[order(-fatal$FATALITIES), ][1:8, ]
# Listing events with highest injuries
injury8 <- injury[order(-injury$INJURIES), ][1:8, ]
par(mfrow = c(1, 2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(fatal8$FATALITIES, las = 3, names.arg = fatal8$EVTYPE, main = "Events with Highest Fatalities", 
        ylab = "Number of fatalities", col = "light green")
barplot(injury8$INJURIES, las = 3, names.arg = injury8$EVTYPE, main = "Events with Highest Injuries", 
        ylab = "Number of injuries", col = "light green")
```
##Plotting events with highest Property damage and highest crop damage.

Highest Property damage and highest crop damage for Top 8 events were calculated. For better understanding and comparision these values were plotted as follows.

```{r}
# Finding events with highest property damage
propdmg8 <- propdmg[order(-propdmg$PROPDMGVAL), ][1:8, ]
# Finding events with highest crop damage
cropdmg8 <- cropdmg[order(-cropdmg$CROPDMGVAL), ][1:8, ]
par(mfrow = c(1, 2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(propdmg8$PROPDMGVAL/(10^9), las = 3, names.arg = propdmg8$EVTYPE, 
        main = "Events with Highest Property Damages", ylab = "Damage Cost ($ billions)", 
        col = "lightgreen")
barplot(cropdmg8$CROPDMGVAL/(10^9), las = 3, names.arg = cropdmg8$EVTYPE, 
        main = "Events With Highest Crop Damages", ylab = "Damage Cost ($ billions)", 
        col = "lightgreen")
```
##Results and Conclusions
The maximum number of fatalities and injuries are caused by Tornados. Excessive Heat follow for fatalities and Thunderstorm wind for injuries.

Floods caused the maximum property damage where as Drought caused the maximum crop damage. Second major events that caused the maximum damage was Hurricanes/Typhoos for property damage and Floods for crop damage.
## Antonio Quintieri, 18/01/18

